---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# drivesR

<!-- badges: start -->
<!-- badges: end -->

The goal of drivesR is to facilitate data management and analysis for members of the [Diverse Rotations Improve Valuable Ecosystems Services (DRIVES) Project](https://www.drives-network.org/) and anyone interested in using the publicly-available portions of the DRIVES database. 

The DRIVES database contains yield and weather data from 21 long-term agricultural field experiments with a crop rotation component. You can find out more about our research and participating sites at our website, [drives-network.org/](https://www.drives-network.org/). Metadata and a full data inventory can be found at the [DRIVES database repository](https://hdl.handle.net/10779/USDA.ADC.28654694) on Ag Data Commons: https://hdl.handle.net/10779/USDA.ADC.28654694.

As of May 2025, access to DRIVES database requires an API key. Project members and collaborators can request an API key from the DRIVES data manager. Other users must complete our [data use survey](https://www.drives-network.org/data-request) and agree to our data use terms. Outside users will have read access to a subset of the data excluding sites and site years embargoed from public access. 


## Installation

You can install the development version of drivesR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("DRIVES-Project/drivesR")
```

## Example

Detailed instructions and examples can be found in the vignette: 
```{r example}
library(drivesR)
vignette("drivesR-data-use")

```
